Drug sensitivity prediction with the normal inverse Gaussian prior informed by external data

### 2019-11-22 docs/manuscript1.Rnw
fixed error in simulation_gdsc.R simulation 3

	The normal inverse Gaussian (NIG) prior model was introduced in 
  \cite{barndorff-nielsen_hyperbolic_1978} and since 
  \cite{barndorff-nielsen_normal_1997} it is routinely applied in mathematical 
  finance (see, e.g., \cite{kalemanova_normal_2007}). Here we extend it with an
  additional global variance component $\tau_d^2$. Supplementary Material
  (SM) Section \ref{sm-sec:prior} contains more details on the NIG prior.
  To illustrate the effect of the NIG prior on the posterior mean, we consider 
  the prior reparametrised as in 
  \cite{carvalho_handling_2009}, i.e., in terms of shrinkage weights 
  $\kappa_{jd}=1/(1 + \gamma_{jd}^2) \in (0, 1)$.
  Under the (simplified) normal means model, i.e., 
  $\X= \begin{bmatrix} \x_1 & \cdots \x_n \end{bmatrix} \tr =\I_p$, with fixed 
  $\tau_d^2=\sigma_d^2=1$, the resulting conditional
  posterior mean for the $\beta_{jd}$ is
  $\E(\beta_{jd} | y_{jd}, \kappa_{jd}) = (1 - \kappa_{jd}) y_{jd}$. 
  Thus, $\kappa_{jd}=0$ implies no shrinkage of $\beta_{jd}$ and $\kappa_{jd}=1$
  implies full shrinkage towards zero. Figure 
  \ref{fig:dens_kappa} depicts the prior on $\kappa_{jd}$ implied by 
  several choices of $\beta_{jd}$ prior.
<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors. Different line types correspond to different hyperparameter settings. The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
@

@@

	\section{Simulations}
	\subsection{Fixed drug component}
	This section investigates the empirical 
	Bayes estimation properties of the model in a simulated setting; its main aim 
	is to assess hyperparameter recovery. It is a data based simulation,
	wherein the responses are simulated from a synthetic model, but the
	features are taken from the real GDSC data introducted in Section 
	\ref{sec:gdsc}. 
	The real GDSC features contain strong collinearities. Such strong
	collinearities in the design matrix impede correct parameter estimation. 
	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
	with the more modest aim of approximately correct estimation of the 
	hyperparameters.

	A pre-processing step, as explained in Section \ref{sec:gdsc}, selects around 
	300 features per drug, while the 168 drug sensitivities for 482 cell lines are 
	simulated from model (\ref{eq:likelihood}) and 
	(\ref{eq:prior}), with two adjustments: we fix $\forall d: \tau_d^2=1$ 
	and draw the error variances as $\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
	such that the prior $\sigma_d^2$ mean and variance are both one.
	We create four external dummy feature covariates that code for four 
	randomly assigned groups of features. We set $\bm{\alpha}_{\text{feat}}$ such
	that the $\gamma_{jd}^2$ of the four groups of features have prior means
	$\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$ 
	($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$). 
	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. We estimate 
	four models: (i) the NIG model as in Section \ref{sec:estimation}, (ii) an 
	NIG model where we fix $\lambda_{\text{feat}}$ to its true value, and two NIG
	models without the inclusion of the external covariates, again with 
	$\lambda_{\text{feat}}$ either (iii) estimated or (iv) fixed to its true 
	value. Exclusion of the external covariates as in models (iii) and (iv) 
	amounts to direct estimation of one common expected 
	prior mean $\phi$, instead of regression estimates for the $\phi_{jd}$ as
	in models (i) and (ii). All 
	models are fit on 241 randomly selected cell lines, while performance measures 
	are estimated from the remaining 241 cell lines.

<<include=FALSE>>=
load("results/simulations_gdsc_set1.Rdata")
mapmse <- apply(res[, c(7:12)], 2, median)
mapmset <- apply(res[, c(13:18)], 2, median)
melbo <- apply(elbo, 2, median)
phi <- 1/(set$alphaf %*% t(unname(model.matrix(~ factor(1:4)))))
mphi <- apply(1/est[, c(11, 13)], 2, median)
mlambdaf <- apply(est[, c(5, 12)], 2, median)
@	
	Figure \ref{fig:simulations_gdsc_est1} shows the estimated
	$\bm{\alpha}_{\text{feat}}$ together with its true value for models (i) and 
	(ii). The figure shows that $\bm{\alpha}_{\text{feat}}$ is consistently
	underestimated 
	in the full NIG model (both with estimated and fixed $\lambda_{\text{feat}}$),
	while $\lambda_{\text{feat}}$ is overestimated as 
	\Sexpr{round(mlambdaf[1], 2)} in model (i) (the true value is 1). 
	This translates into
	underestimated prior means $\phi_{jd}$ as presented in Figure 
	\ref{fig:simulations_gdsc_phi1}, which in turn implies underestimated 
	$\beta_{jd}$ variances. The overestimated $\lambda_{\text{feat}}$ translates 
	into underestimated $\gamma_{jd}^2$ variances. Underestimation of variances is 
	a common phenomenon in variational Bayes approximations. In addition, 
	$\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ are higher level 
	parameters, which are, in general, hard to estimate. We note 
	that even though $\bm{\alpha}_{\text{feat}}$ and thus $\phi_{jd}$ are 
	estimated with bias, the orders of their elements are correct.
<<simulations_gdsc_est1, fig.cap="Estimated hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>= 
@

<<simulations_gdsc_phi1, fig.cap="Estimated prior means $\\bm{\\phi}_{jd}$ for the four groups of features from simulations with (a) estimated and (b) fixed $\\lambda_{\\text{feat}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1/2>>= 
@

	The models (iii) and (iv) give median $\phi$ estimates of
	\Sexpr{round(mphi[1], 2)} and \Sexpr{round(mphi[2], 2)}, 
	respectively. The $\phi$ estimates are
	underestimated, compared to true $\phi$ mean \Sexpr{round(mean(phi), 2)}. 
	Model (iii) slightly underestimates the median
	$\lambda_{\text{feat}}$ at \Sexpr{round(mlambdaf, 2)[2]} 
	(the true value is one) and thus overestimates the 
	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
	prior mean 
	$\phi$, that is not able to capture part of the variation in the true feature
	specific $\phi_{jd}$. The median evidence 
	lower bounds (ELBO), which approximate the marginal likelihoods of models
	(i)-(iv), are estimated as \Sexpr{round(melbo, 2)[1:3]}, and 
	\Sexpr{round(melbo, 2)[4]}, respectively. This implies that model (i),
	with estimated $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ fits
	the data best (the log Bayes factors comparing models (ii)-(iv) to model (i) 
	are \Sexpr{round(melbo[-c(1, 4)] - melbo[1], 2)}, and 
	\Sexpr{round(melbo[4] - melbo[1], 2)}, respectively, and thus highly favouring
	model (i)). The prediction MSEs for the four models on the test 
	data and gaps between training and test prediction MSE are displayed in Table
	\ref{tab:pmse1}. Table \ref{tab:pmse1} confirms that the models that
	include the external data (models (i) and (ii)) learn the underlying structure 
	in the data better than models (iii) and (iv) that do not include the external 
	data.
<<pmse1>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
load("results/simulations_gdsc_set1.Rdata")
mapmse <- apply(res[, c(7:12)], 2, median)
mapmset <- apply(res[, c(13:18)], 2, median)
tab <- rbind(mapmse, mapmse - mapmset)[, -c(5, 6)]
rownames(tab) <- c("$\\text{MSE}_{\\text{test}}$",
                  "$\\text{MSE}_{\\text{test}} - \\text{MSE}_{\\text{train}}$")
options(knitr.kable.NA="", digits=3)
kable(tab,
      col.names=c("(i)", "(ii)", "(iii)", "(iv)"),
      caption="Median prediction MSEs for the models (i)-(iv) in the fixed 
      $\\tau_d^2$ simulation setting.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

  \subsection{Random drug component}
	To investigate the interplay of the feature and drug variance components, we 
	repeat the whole simulation study with random $\tau_d^2$. Similarly as with 
	the $\gamma_{jd}^2$, we randomly create four groups of drugs and corresponding
	external drug dummy covariates. We simulate the $\tau_d^2$ as in
	(\ref{eq:tauprior}), where we set $\bm{\alpha}_{\text{drug}} = 
	\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$, such
	that the $\tau_{d}^2$ of the four groups of features have prior means
	$\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The scale $\lambda_{\text{drug}}$ is 
	set to one. We estimate the same four models as in the fixed $\tau_d^2$ 
	setting, but in addition estimate the $\tau_d^2$ priors as well. Again, we
	use two different versions: one where we estimate 
	$\bm{\lambda}$ and one where we fix it to its true value.

<<include=FALSE>>=
load("results/simulations_gdsc_set2.Rdata")
mapmse <- apply(res[, c(7:12)], 2, median)
mapmset <- apply(res[, c(13:18)], 2, median)
melbo <- apply(elbo, 2, median)
phi <- 1/(set$alphaf %*% t(unname(model.matrix(~ factor(1:4)))))
chi <- 1/(set$alphad %*% t(unname(model.matrix(~ factor(1:4)))))
mphi <- apply(1/est[, c(21, 25)], 2, median)
mchi <- apply(1/est[, c(22, 26)], 2, median)
mlambdaf <- apply(est[, c(9, 23)], 2, median)
mlambdad <- apply(est[, c(10, 24)], 2, median)
@	
	Figures \ref{fig:simulations_gdsc_est2} and 
	\ref{fig:simulations_gdsc_phi2} display the estimated $\bm{\alpha}$, 
	$\phi_{jd}$, and $\chi_d$ for models (i) and (ii). The scale parameters were
	estimated as $\lambda_{\text{feat}}=\Sexpr{round(mlambdaf[1], 2)}$ and
	$\lambda_{\text{drug}}=\Sexpr{round(mlambdad[1], 2)}$ in model (i). From 
	Figure \ref{fig:simulations_gdsc_est2} we see that estimation of
	$\bm{\alpha}_{\text{feat}}$ is again biased. However, from 
	Figure \ref{fig:simulations_gdsc_phi2} we see that the scale and ordering of
	the $\phi_{jd}$ are correclty estimated. We conjecture that the drop in 
	estimation accuracy compared to the previous setting stems from the
	difficult estimation of the $\tau_d^2$ prior. Estimation of this prior 
	depends on only 168 observed drugs, while the $\gamma_{jd}^2$ prior is 
	is estimated from approximately $300 \cdot 168=50400$ features. 
	\cite{wiel_learning_2019} show in a simple regression setting that if the 
	number of samples and the number of parameters on which the EB is based are of
	the same order, empirical Bayes performs worse than when one of the two is
	large compared to the other. 
	Here, the number of samples $n=241$ and $D=168$ are of the same order, leading
	to a suboptimal EB estimation performance.

<<simulations_gdsc_est2, fig.cap="Estimated hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and $\\bm{\\alpha}_{\\text{drug}}$ from simulations with (a) \\& (c) estimated and (b) \\& (d) fixed $\\lambda_{\\text{feat}}$ and $\\lambda_{\\text{drug}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1>>= 
@

<<simulations_gdsc_phi2, fig.cap="Estimated prior means $\\bm{\\phi}_{jd}$ and $\\bm{\\chi}_{d}$ for the four groups of features and drugs from simulations with (a) \\& (c) estimated and (b) \\& (d) fixed $\\lambda_{\\text{feat}}$ and $\\lambda_{\\text{drug}}$ (models (i) and (ii), respectively).", out.width="100%", fig.asp=1>>= 
@

  We argue however, that even though the prior parameters are not estimated
  correctly, there is still some gain in predictive performance.
  Prediction MSEs on the test data the training-test gaps in prediction MSE are 
  shown in Table \ref{tab:pmse2}. The comparisons of models (i) and
	(ii) vesus models (iii) and (iv) confirm that even though EB estimation is
	suboptimal, the external data do add something in predictive performance. 
<<pmse2>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
load("results/simulations_gdsc_set2.Rdata")
mapmse <- apply(res[, c(7:12)], 2, median)
mapmset <- apply(res[, c(13:18)], 2, median)
tab <- rbind(mapmse, mapmse - mapmset)[, -c(5, 6)]
rownames(tab) <- c("$\\text{MSE}_{\\text{test}}$",
                  "$\\text{MSE}_{\\text{test}} - \\text{MSE}_{\\text{train}}$")
options(knitr.kable.NA="", digits=3)
kable(tab,
      col.names=c("(i)", "(ii)", "(iii)", "(iv)"),
      caption="Median prediction MSEs for the models (i)-(iv) in the random 
      $\\tau_d^2$ simulation setting.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

@@

  From the $\bm{\alpha}$ estimates in Tables \ref{tab:results_drug} and
  \ref{tab:results_feat} we see that there
  seems to be an effect of the
  external variables. If we look at the NIG1 model, that includes all external 
  covariates, we see that approved drugs are easier to predict, due to small
  prior precision, followed by drugs that are in clinical development,
  as we expected. The cytotoxic drugs are the hardest to predict, 
  with the largest prior precision (smallest prior variance), followed by the 
  targeted drugs and the unknown drugs. This is again according to expectation.
  For the unkown drugs, we are unable to confirm the small precision. Genes 
  that are in the target pathway have 
  smaller prior precisions, according to expectation, which confirms that these
  genes are more important for prediction of drug sensitivity. The estimates 
  from
  the bSEM model are also according to expectation: genes that are in the target
  pathway have a much smaller precision than genes that are not in the target
  pathway. The 
  effects of the pathway indicator seem to be much larger in the bSEM model. 
<<results_drug>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
load("results/data_gdsc_res1.Rdata")
tab1 <- cbind(tab[, 1], c(NA, 0, NA), tab[, c(3, 2)], c(NA, 0, NA), 
              tab[, c(4, 5)])
options(knitr.kable.NA="", digits=3)
kable(tab1,
      col.names=c("intercept", "approved", "development", 
                  "experimental", "targeted", "unkown", "cytotoxic"),
      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<results_feat>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
load("results/data_gdsc_res1.Rdata")
tab2 <- cbind(tab[, 6], c(NA, 0, 0), tab[, 7])
options(knitr.kable.NA="", digits=3)
kable(tab2,
      col.names=c("intercept", "not in pathway", "in pathway"),
      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

  The median prediction MSEs and the median ELBOs, calculated on the test data
  are displayed in Table \ref{tab:results_data}. In terms of 
  is lowest in the NIG models 
  (\Sexpr{round(mapmse[2], 2)} and \Sexpr{round(mapmse[3], 2)}, in the
  NIG1 and NIG2 models, respectively). The bSEM model is not able to 
  improve on the null model with an median prediction MSE of 
  \Sexpr{round(mapmse[4], 2)} compared to the null model's MSE of 
  \Sexpr{round(mapmse[1], 2)}.
<<results_data>>= 
load("results/data_gdsc_res1.Rdata")
mapmse <- apply(apmse, 2, median)
melbo <- apply(elbo, 2, median)
tab <- rbind(mapmse, c(melbo, NA, NA))
rownames(tab) <- c("MSE", "ELBO")
options(knitr.kable.NA="", digits=3)
kable(tab,
      col.names=c("NIG1", "NIG2", "bSEM", "lasso", "ridge"),
      caption="Median prediction MSE and median ELBO estimated on the test data", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

>>>>

	The normal inverse Gaussian (NIG) prior model was introduced in 
  \cite{barndorff-nielsen_hyperbolic_1978} and since 
  \cite{barndorff-nielsen_normal_1997} it is routinely applied in mathematical 
  finance (see, e.g., \cite{kalemanova_normal_2007}). Here we extend it with an
  additional global variance component $\tau_d^2$. Supplementary Material
  (SM) Section \ref{sm-sec:prior} contains more details on the NIG prior.
  To illustrate the effect of the NIG prior on the posterior mean, we consider 
  the prior reparametrised as in 
  \cite{carvalho_handling_2009}, i.e., in terms of shrinkage weights 
  $\kappa_{jd}=1/(1 + \gamma_{jd}^2) \in (0, 1)$.
  Under the (simplified) normal means model, i.e., 
  $\X= \begin{bmatrix} \x_1 & \cdots \x_n \end{bmatrix} \tr =\I_p$, with fixed 
  $\tau_d^2=\sigma_d^2=1$, the resulting conditional
  posterior mean for the $\beta_{jd}$ is
  $\E(\beta_{jd} | y_{jd}, \kappa_{jd}) = (1 - \kappa_{jd}) y_{jd}$. 
  Thus, $\kappa_{jd}=0$ implies no shrinkage of $\beta_{jd}$ and $\kappa_{jd}=1$
  implies full shrinkage towards zero. Figure 
  \ref{fig:dens_kappa} depicts the prior on $\kappa_{jd}$ implied by 
  several choices of $\beta_{jd}$ prior.

<<dens_kappa, fig.cap="Implied prior densities $\\pi(\\kappa_{jd})$ for the (a) NIG, (b) Student's $t$, and (c) lasso priors. Different line types correspond to different hyperparameter settings. The hyperparameter settings were chosen to show some possible, distinct shapes that each of the priors can take.", out.width="100%", fig.asp=1/3>>=
@

@@

	\section{Simulations}
	\subsection{Setup}
	This section investigates the empirical 
	Bayes estimation properties of the model in a simulated setting; its main aim 
	is to assess hyperparameter recovery. It is a data based simulation,
	wherein the responses are simulated from a synthetic model, but the
	features are taken from the real GDSC data introducted in Section 
	\ref{sec:gdsc}. 
	The real GDSC features contain strong collinearities. Such strong
	collinearities in the design matrix impede correct parameter estimation. 
	We therefore replace the ambition of correctly estimating the $\beta_{jd}$
	with the more modest aim of approximately correct estimation of the 
	hyperparameters.

	A pre-processing step selects 100 features with largest variance, while the 
	168 drug sensitivities for 482 cell lines are 
	simulated from three versions of model (\ref{eq:likelihood}) and 
	(\ref{eq:prior}). All versions draw the error variances as 
	$\forall d: \sigma_d^2 \sim \Gamma^{-1}(3,2)$,
	such that the prior $\sigma_d^2$ mean and variance are both one. 
	Version 1 fixes $\forall d: \tau_d^2=1$, version 2 fixes
	$\forall j,d: \gamma_{jd}^2=1$, while version 3 draws both the $\tau_d^2$ and
	$\gamma_{jd}^2$ according to (\ref{eq:prior}). In version 1, we
	create four external dummy feature covariates that code for four 
	randomly assigned approximately equally sized groups of features. We set
	$\bm{\alpha}_{\text{feat}}$ such
	that the $\gamma_{jd}^2$ of the four groups of features have prior means
	$\phi_{jd} \in \{ 1, 1/2, 1/4, 1/8 \}$ 
	($\bm{\alpha}_{\text{feat}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$). 
	The prior scale parameter is set to $\lambda_{\text{feat}}=1$. Version 2 
	follows a similar procedure for the $\tau_d^2$. We randomly create four
	groups of drugs with corresponding external drug dummy variables and set
	$\bm{\alpha}_{\text{drug}}=\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$,
	such that we have $\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The
	scale is set to $\lambda_{\text{drug}}=1$. Version 3 combines the procedures
	from versions 1 and 2 to draw the $\gamma_{jd}^2$ and $\tau_d^2$.

	We estimate four models: (i) an NIG model that does include external 
	covariates in the estimation, (ii) the NIG model estimated as in Section 
	\ref{sec:estimation}, and regular (iii) lasso and (iv) ridge models.
	Exclusion of the external covariates as in model (i)
	amounts to direct estimation of one common expected 
	prior means $\phi$ and/or $\chi$, instead of regression estimates for the
	$\phi_{jd}$ and/or $\chi_d$ as in model (ii). All 
	models are fit on 241 randomly selected cell lines, while performance measures 
	are estimated from the remaining 241 cell lines.

	\subsection{Results}
<<simulation_gdsc>>=
res1 <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
res2 <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
res3 <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
temp1 <- res1[, 1]
temp2 <- res2[, 1]
temp3 <- res3[, 1]
res1 <- as.matrix(res1[, -1])
res2 <- as.matrix(res2[, -1])
res3 <- as.matrix(res3[, -1])
rownames(res1) <- temp1
rownames(res2) <- temp2
rownames(res3) <- temp3
lambda <- apply(res1[rownames(res1)=="lambdaf", c(1, 2)], 2, median)
apply(res2[rownames(res2)=="lambdad", c(1, 2)], 2, median)
apply(res3[rownames(res3)=="lambdaf", c(1, 2)], 2, median)
apply(res3[rownames(res3)=="lambdad", c(1, 2)], 2, median)
est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
melbo <- apply(res[rownames(res)=="elbo", c(1, 2)], 2, median)
melbot <- apply(res[rownames(res)=="elbot", c(1, 2)], 2, median)

mphi <- median(1/res[rownames(res)=="alphaf0", 1])
mlambdaf <- median(1/res[rownames(res)=="lambdaf", 1])
@	
	Figure \ref{fig:simulations_gdsc_est1} shows the estimated
	$\bm{\alpha}_{\text{feat}}$ together with its true value for model
	(ii) in version 1 of the simulation study. Figure 
	\ref{fig:simulations_gdsc_est1}a shows that 
	$\bm{\alpha}_{\text{feat}}$ is slightly but 
	consistently underestimated. On the $\phi_{jd}$ scale this underestimation is
	almost undetectable as depicted in Figure
	\ref{fig:simulations_gdsc_est1}b. $\lambda_{\text{feat}}$ is slightly 
	overestimated with median estimate \Sexpr{round(lambda[2], 2)} (the
	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
	into underestimated $\gamma_{jd}^2$ variances. Underestimation of variances is 
	a common phenomenon in variational Bayes approximations. In addition, 
	$\lambda_{\text{feat}}$ is a higher level scale
	parameter, which are, in general, hard to estimate. 
<<simulations_gdsc_est1, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulations with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>= 
@

	Model (i) gives a median $\phi$ estimate of \Sexpr{round(est.phi, 2)},  
	equal to the true mean of the $\phi_{jd}$, \Sexpr{round(true.phi, 2)}. 
	$\lambda_{\text{feat}}$ is slightly underestimated with median estimate
	\Sexpr{round(lambda[1], 2)} (the true value is one) and thus overestimates the 
	prior $\gamma_{jd}^2$ variance. This is likely to compensate for the common
	prior mean $\phi$, that is not able to capture part of the variation in the 
	true feature specific $\phi_{jd}$.

	The median evidence 
	lower bounds (ELBO), which approximate the marginal likelihoods of models
	(i)-(iv), are estimated as \Sexpr{round(melbo, 2)[1:3]}, and 
	\Sexpr{round(melbo, 2)[4]}, respectively. This implies that model (i),
	with estimated $\bm{\alpha}_{\text{feat}}$ and $\lambda_{\text{feat}}$ fits
	the data best (the log Bayes factors comparing models (ii)-(iv) to model (i) 
	are \Sexpr{round(melbo[-c(1, 4)] - melbo[1], 2)}, and 
	\Sexpr{round(melbo[4] - melbo[1], 2)}, respectively, and thus highly favouring
	model (i)). The prediction MSEs for the four models on the test 
	data and gaps between training and test prediction MSE are displayed in Table
	\ref{tab:mse1}. Table \ref{tab:mse1} confirms that the models that
	include the external data (models (i) and (ii)) learn the underlying structure 
	in the data better than models (iii) and (iv) that do not include the external 
	data.
<<mse1>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp
memse <- apply(res[rownames(res)=="emse", ], 2, median)
mpmse <- apply(res[rownames(res)=="pmse", ], 2, median)
mpmset <- apply(res[rownames(res)=="pmset", ], 2, median)
tab <- rbind(memse, mpmse, mpmse - mpmset)
rownames(tab) <- c("$\\text{EMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}}$",
                   "$\\text{PMSE}_{\\text{test}} - \\text{PMSE}_{\\text{train}}$")
options(knitr.kable.NA="", digits=3)
kable(tab,
      col.names=colnames(tab),
      caption="Median estimation MSE, prediction MSEs, and difference between 
      prediction MSEs on the test and training data in the fixed 
      $\\tau_d^2$ simulation setting.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

	To investigate the interplay of the feature and drug variance components, we
	repeat the whole simulation study with random $\tau_d^2$. Similarly as with
	the $\gamma_{jd}^2$, we randomly create four groups of drugs and corresponding
	external drug dummy covariates. We simulate the $\tau_d^2$ as in
	(\ref{eq:tauprior}), where we set $\bm{\alpha}_{\text{drug}} =
	\begin{bmatrix} 1 & 1 & 3 & 7 \end{bmatrix} \tr$, such
	that the $\tau_{d}^2$ of the four groups of features have prior means
	$\chi_{d} \in \{ 1, 1/2, 1/4, 1/8 \}$. The scale $\lambda_{\text{drug}}$ is
	set to one. We estimate the same four models as in the fixed $\tau_d^2$
	setting, but in addition estimate the $\tau_d^2$ priors as well. Again, we
	use two different versions: one where we estimate
	$\bm{\lambda}$ and one where we fix it to its true value.

	Figures \ref{fig:simulations_gdsc_est2} displays the estimated $\bm{\alpha}$,
	$\phi_{jd}$, and $\chi_d$ for models (i) and (ii). The scale parameters were
	estimated as $\lambda_{\text{feat}}=\Sexpr{round(rnorm(1), 2)}$ and
	$\lambda_{\text{drug}}=\Sexpr{round(rnorm(1)[1], 2)}$ in model (i). From
	Figure \ref{fig:simulations_gdsc_est2} we see that estimation of
	$\bm{\alpha}_{\text{feat}}$ is again biased. However, from
	Figure \ref{fig:simulations_gdsc_phi2} we see that the scale and ordering of
	the $\phi_{jd}$ are correclty estimated. We conjecture that the drop in
	estimation accuracy compared to the previous setting stems from the
	difficult estimation of the $\tau_d^2$ prior. Estimation of this prior
	depends on only 168 observed drugs, while the $\gamma_{jd}^2$ prior is
	is estimated from approximately $300 \cdot 168=50400$ features.
	\cite{wiel_learning_2019} show in a simple regression setting that if the
	number of samples and the number of parameters on which the EB is based are of
	the same order, empirical Bayes performs worse than when one of the two is
	large compared to the other.
	Here, the number of samples $n=241$ and $D=168$ are of the same order, leading
	to a suboptimal EB estimation performance.
<<simulations_gdsc_est2, fig.cap="Estimated (a) hyperparameters $\\bm{\\alpha}_{\\text{feat}}$ and (b) prior means $\\bm{\\phi}_{jd}$ from simulations with fixed $\\tau_d^2$.", out.width="100%", fig.asp=1/2>>= 
@

  We argue however, that even though the prior parameters are not estimated
  correctly, there is still some gain in predictive performance.
  Prediction MSEs on the test data the training-test gaps in prediction MSE are
  shown in Table \ref{tab:pmse2}. The comparisons of models (i) and
	(ii) vesus models (iii) and (iv) confirm that even though EB estimation is
	suboptimal, the external data do add something in predictive performance.

@@

  From the $\bm{\alpha}$ estimates in Tables \ref{tab:results_drug} and
  \ref{tab:results_feat} we see that there
  seems to be an effect of the
  external variables. If we look at the NIG1 model, that includes all external 
  covariates, we see that approved drugs are easier to predict, due to small
  prior precision, followed by drugs that are in clinical development,
  as we expected. The cytotoxic drugs are the hardest to predict, 
  with the largest prior precision (smallest prior variance), followed by the 
  targeted drugs and the unknown drugs. This is again according to expectation.
  For the unkown drugs, we are unable to confirm the small precision. Genes 
  that are in the target pathway have 
  smaller prior precisions, according to expectation, which confirms that these
  genes are more important for prediction of drug sensitivity. The estimates 
  from
  the bSEM model are also according to expectation: genes that are in the target
  pathway have a much smaller precision than genes that are not in the target
  pathway. The 
  effects of the pathway indicator seem to be much larger in the bSEM model. 
<<results_drug>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
# load("results/data_gdsc_res1.Rdata")
# tab1 <- cbind(tab[, 1], c(NA, 0, NA), tab[, c(3, 2)], c(NA, 0, NA), 
#               tab[, c(4, 5)])
# options(knitr.kable.NA="", digits=3)
# kable(tab1,
#       col.names=c("intercept", "approved", "development", 
#                   "experimental", "targeted", "unkown", "cytotoxic"),
#       caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.", 
#       format="latex", booktabs=TRUE, escape=FALSE) %>%
#   kable_styling(latex_options=c("HOLD_position"))
kable(matrix(1, nrow=1, ncol=2),
      caption="$\\bm{\\alpha}_{\\text{drug}}$ estimates from GDSC data.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<results_feat>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
# load("results/data_gdsc_res1.Rdata")
# tab2 <- cbind(tab[, 6], c(NA, 0, 0), tab[, 7])
# options(knitr.kable.NA="", digits=3)
# kable(tab2,
#       col.names=c("intercept", "not in pathway", "in pathway"),
#       caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.", 
#       format="latex", booktabs=TRUE, escape=FALSE) %>%
#   kable_styling(latex_options=c("HOLD_position"))
kable(matrix(1, nrow=1, ncol=2),
      caption="$\\bm{\\alpha}_{\\text{feat}}$ estimates from GDSC data.", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

  The median prediction MSEs and the median ELBOs, calculated on the test data
  are displayed in Table \ref{tab:results_data}. In terms of 
  is lowest in the NIG models 
  (\Sexpr{round(rnorm(1))} and \Sexpr{rnorm(1)}, in the
  NIG1 and NIG2 models, respectively). The bSEM model is not able to 
  improve on the null model with an median prediction MSE of 
  \Sexpr{round(rnorm(1))} compared to the null model's MSE of 
  \Sexpr{round(rnorm(1))}.
<<results_data>>= 
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
# load("results/data_gdsc_res1.Rdata")
# mapmse <- apply(apmse, 2, median)
# melbo <- apply(elbo, 2, median)
# tab <- rbind(mapmse, c(melbo, NA, NA))
# rownames(tab) <- c("MSE", "ELBO")
# options(knitr.kable.NA="", digits=3)
# kable(tab,
#       col.names=c("NIG1", "NIG2", "bSEM", "lasso", "ridge"),
#       caption="Median prediction MSE and median ELBO estimated on the test data", 
#       format="latex", booktabs=TRUE, escape=FALSE) %>%
#   kable_styling(latex_options=c("HOLD_position"))
options(knitr.kable.NA="", digits=3)
kable(matrix(1, nrow=1, ncol=2),
      caption="Median prediction MSE and median ELBO estimated on the test data", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@
